---
title: "R Notebook"
output: html_notebook
---



```{r}
library(caroline)
library(Seurat)
library(stringr)
library(ggplot2)
library(patchwork)

#setwd("D:/scRNA_vasculature/mouse_development/")
setwd("~/Desktop/mouse_development")
```


## ====> Data Acquisition and Preprocessing
# Load GSE104323 from the NCBI GEO database.
# pre-processing: normalization, scaling, and dimensionality reduction

```{r}
data <- read.tab("/mouse_development_data/GSE104323_10X_expression_data.tab",header = T)
row.names(data) <- data[,1]
data <- data[,-1]

meta <- read.table("GSE104323_metadata_barcodes_24185cells.txt", sep = "\t",header = T)[1:24185,]
meta$Sample.name..24185.single.cells. <- paste("X",meta$Sample.name..24185.single.cells., sep="")
meta$Sample.name..24185.single.cells. <- str_replace_all(meta$Sample.name..24185.single.cells., "-",".")
row.names(meta) <- meta$Sample.name..24185.single.cells.
metadata <- meta[,c(5,7)]
```

```{r}
seu_obj <- CreateSeuratObject(counts = data, min.cells = 3, meta.data = metadata)

seu_data <- NormalizeData(object = seu_obj, scale.factor = 1e6)
seu_data <- ScaleData(object = seu_data)
seu_data <- FindVariableFeatures(object = seu_data)
seu_data <- RunPCA(seu_data, verbose = FALSE)
seu_data <- FindNeighbors(seu_data, dims = 1:30)
seu_data <- FindClusters(seu_data, resolution = 1.0, verbose = FALSE)
seu_data <- RunUMAP(seu_data, dims = 1:30)
```

```{r}
seu_data[["percent.mt"]] <- PercentageFeatureSet(seu_data, pattern = "^MT-")
VlnPlot(seu_data, features = c("nFeature_RNA"), group.by = 'characteristics..cell.cluster') + theme(axis.text.x = element_text(size = 9), legend.text = element_text(size = 9))
VlnPlot(seu_data, features = c("nCount_RNA"), group.by = 'characteristics..cell.cluster') + theme(axis.text.x = element_text(size = 9), legend.text = element_text(size = 9)) 
FeatureScatter(seu_data, feature1 = "nCount_RNA", group.by = 'characteristics..cell.cluster', feature2 = "nFeature_RNA")
```

```{r}
DimPlot(seu_data, reduction = "umap",  label = TRUE)
DimPlot(seu_data, group.by = 'characteristics..cell.cluster') + coord_fixed(ratio = 0.6)
DimPlot(seu_data, group.by = 'characteristics..cell.cluster',  label = TRUE) + coord_fixed(ratio = 0.6)

```
```{r}
Idents(seu_data) <- seu_data@meta.data$characteristics..age

# RenameIdnets
seu_data <- RenameIdents(seu_data, 'E16.5' = "embryonic")
seu_data <- RenameIdents(seu_data, 'P0' = "postnatal", 'P5' = "postnatal")
seu_data <- RenameIdents(seu_data, 'P18' = "weaning",'P19' = "weaning",  'P23' = "weaning")
seu_data <- RenameIdents(seu_data, 'P120' = "adult", 'P132' = "adult")
Idents(seu_data) <- factor(Idents(seu_data), levels = c("embryonic", "postnatal", "weaning","adult"))
```



## ====> check out cell type markers
```{r}
all.markers <- FindAllMarkers(object = seu_data, verbose = FALSE, logfc.threshold = 0.2)
```





## ====> check out gpr 4 briefly

```{r}
my_levels <- c("E16.5", "P0", "P5", "P18", "P19", "P23", "P120", "P132")
seu_data@meta.data[["characteristics..age"]] <- factor(x = seu_data@meta.data[["characteristics..age"]], levels = my_levels)
VlnPlot(seu_data, features = "Gpr4", group.by = "characteristics..age") + labs(title = "Gpr4 expression level from E16.5 to P132")
```

```{r}
FeaturePlot(seu_data, features = "Gpr4")
```

# Gpr4 VlnPlot across cell types
```{r}
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "RGL"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "RGL")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Astro-adult"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Astro-adult")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Astro-juv"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Astro-juv")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Immature-Astro"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Immature-Astro")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "RGL_young"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "RGL_youn")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Ependymal"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Ependymal")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "PVM"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "PVM")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "MiCajal-Retziusoglia"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "MiCajal-Retziusogli")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Endothelial"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Endothelial")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "VLMC"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "VLMC")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "MOL"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "MOL")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "NFOL"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "NFOL")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "OPC"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "OPC")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "nIPC-perin"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "nIPC-perin")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "nIPC"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "nIPC")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Neuroblast"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Neuroblast")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Immature-GC"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Immature-GC")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Immature-Pyr"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Immature-Pyr")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Cajal-Retzius"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Cajal-Retzius")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Immature-GABA"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Immature-GABA")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "GABA"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "GABA")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "CA3-Pyr"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "CA3-Pyr")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "GC-juv"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "GC-juv")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "GC-adult"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "GC-adult")
```

```{r}
VlnPlot(seu_data, features = "H19",group.by = "characteristics..age")

VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Endothelial"), features = "H19",group.by = "characteristics..age") + labs(title = "H19 expression level from E16.5 to P132")
VlnPlot(subset(x = seu_data, subset = characteristics..cell.cluster == "Endothelial"), features = "Gpr4",group.by = "characteristics..age") + labs(title = "Gpr4 expression level from E16.5 to P132")

```

## ====> check out other EC-specific markers
```{r}
ec_markers <- FindMarkers(seu_data, only.pos = TRUE, group.by = "characteristics..cell.cluster", ident.1 = "Endothelial", ident.2 = c('RGL','Astro-adult','Astro-juv','Immature-Astro','RGL_young','Ependymal','PVM','MiCajal-Retziusoglia','VLMC','MOL','NFOL','OPC','nIPC-perin','nIPC','Neuroblast','Immature-GC','Immature-Pyr','Cajal-Retzius','Immature-GABA','GABA','CA3-Pyr','GC-juv','GC-adult'), logfc.threshold = 0.2)
```

# qc the ec markers
```{r}
ec_markers <- ec_markers[ec_markers$pct.1 > 0.6 & ec_markers$pct.2 < 0.1 & ec_markers$p_val < 0.05, ]
write.csv(ec_markers, "result/deg_tab_ec_specific_46.csv")

# get 7xx degs -> 46 degs
```

# plot ec markers
```{r}
#cwd <- "~/Desktop/mouse_development/brain_vasculature_co2"
cwd <- "E:/Shuting/gpr/brain_vasculature_co2-master"

# edit HERE
pdf(file= paste0(cwd, "/result/ec_ov_ratio_adj_46genes.pdf"), onefile = TRUE)
candidates <- row.names(ec_markers[1:46,])

for (i in seq(1, length(candidates), by = 9)){
  p1 <- FeaturePlot(seu_data, features = candidates[i:(i+8)], pt.size = 0.2, ncol = 3)
  #p2 <- VlnPlot(seu_data, features = candidates[i:(i+8)], pt.size = 0.2, group.by = 'characteristics..cell.cluster', ncol = 3)
  print(p1)
  #print(p2)
}
dev.off()
```

#  arrow down the markers
# ---------- NOTE ----------
# from ec_ov_ratio_adj.pdf
# ec-specific markers: 
# c("Cldn5", "Flt1",  "Ly6c1", "Igfbp7", "Itm2a", "Slco1a4", "Ptprb", "Esam", "Adgrl4", "Eng", "Ctla2a", "Itm2a", "Pglyrp1", "Kdr", "Pltp",  "Cd34", "Adgrl4", "BC028528", "Vwa1", "Adgrf5", "Ebf1", "Slc22a8", "Fn1", "St3gal6", "Ccdc141", "Ifitm3", "Klf2")

```{r}
ec_fil_genes <- c("Cldn5", "Flt1",  "Ly6c1", "Igfbp7", "Slco1a4", "Ptprb", "Esam", "Eng", "Ctla2a", "Itm2a", "Pglyrp1", "Kdr", "Pltp", "Cd34", "Adgrl4", "BC028528", "Vwa1", "Adgrf5", "Ebf1", "Slc22a8", "Fn1", "St3gal6", "Ccdc141", "Ifitm3", "Klf2")
ec_markers$gene <- row.names(ec_markers)
ec_fil_markers <- ec_markers[ec_markers$gene %in% ec_fil_genes,]
write.csv(ec_fil_markers, "result/deg_tab_ec_specific_25.csv")
```


```{r}
# edit HERE
pdf(file= paste0(cwd, "/result/ec_ov_ratio_adj_25genes.pdf"), onefile = TRUE)
candidates <- row.names(ec_fil_markers)

for (i in seq(1, length(candidates), by = 4)){
  p1 <- FeaturePlot(seu_data, features = candidates[i:(i+3)], pt.size = 0.2, ncol = 2)
  #p2 <- VlnPlot(seu_data, features = candidates[i:(i+3)], pt.size = 0.2, group.by = 'characteristics..cell.cluster', ncol = 3)
  print(p1)
  #print(p2)
}
dev.off()
```


```{r}
DoHeatmap(object = seu_data, features = row.names(ec_markers), label=FALSE) + theme(axis.text.x = element_text(size = 4), axis.text.y = element_text(size = 5), legend.text = element_text(size = 6))
DoHeatmap(object = seu_data, features = row.names(ec_fil_markers),  label=FALSE) + theme(axis.text.x = element_text(size = 4), axis.text.y = element_text(size = 5), legend.text = element_text(size = 6))

```




## ====> subset the ECs and label the temporal clusters
# NOTE
# EC: 543 cells

```{r}
# subset
seu_data[["cluster_idents"]] <- Idents(seu_data)
Idents(seu_data) <- seu_data@meta.data$characteristics..cell.cluster
endo_cells <- subset(seu_data, idents = c("Endothelial"))
```


```{r}
# count ECs

# clustering
endo_cells <- NormalizeData(object = endo_cells)
endo_cells <- ScaleData(object = endo_cells)
endo_cells <- FindVariableFeatures(object = endo_cells)
endo_cells <- RunPCA(endo_cells, verbose = FALSE)
endo_cells <- FindNeighbors(endo_cells, dims = 1:30)
endo_cells <- FindClusters(endo_cells, resolution = 1, verbose = FALSE)
endo_cells <- RunUMAP(endo_cells, dims = 1:30)

```


# plot
```{r}
DimPlot(endo_cells, reduction = "umap",  label = TRUE)
```

```{r}
Idents(endo_cells) <- endo_cells@meta.data$characteristics..age

# RenameIdnets
endo_cells <- RenameIdents(endo_cells, 'E16.5' = "embryonic")
endo_cells <- RenameIdents(endo_cells, 'P0' = "postnatal", 'P5' = "postnatal")
endo_cells <- RenameIdents(endo_cells, 'P18' = "weaning",'P19' = "weaning",  'P23' = "weaning")
endo_cells <- RenameIdents(endo_cells, 'P120' = "adult", 'P132' = "adult")
Idents(endo_cells) <- factor(Idents(endo_cells), levels = c("embryonic", "postnatal", "weaning","adult"))

DimPlot(endo_cells, reduction = "umap",  label = TRUE)

```

## ====> temporal-specific markers
```{r}
DefaultAssay(endo_cells)  <- "RNA"

deg1 <- FindMarkers(endo_cells, ident.1 = "postnatal", ident.2 = "embryonic", logfc.threshold = 0.2)
write.csv(deg1, "result/deg_tab_ec_postnatal.csv")
deg2 <- FindMarkers(endo_cells, ident.1 = "weaning", ident.2 = "postnatal", logfc.threshold = 0.2)
write.csv(deg2, "result/deg_tab_ec_weaning.csv")
deg3 <- FindMarkers(endo_cells, ident.1 = "adult", ident.2 = "weaning", logfc.threshold = 0.2)
write.csv(deg3, "result/deg_tab_ec_adult.csv")

```

# qc
```{r}
pct_thres <- 0.2
fc_thres <- 0.1
pv <- 0.05

deg1_filtered <- deg1[(deg1$p_val<pv & abs(deg1$avg_log2FC) > fc_thres),]
deg2_filtered <- deg2[(deg2$p_val<pv & abs(deg2$avg_log2FC) > fc_thres),]
deg3_filtered <- deg3[(deg3$p_val<pv & abs(deg3$avg_log2FC) > fc_thres),]


deg1_filtered <- deg1_filtered[(((deg1_filtered$avg_log2FC > fc_thres)&(deg1_filtered$pct.1>pct_thres))|((deg1_filtered$avg_log2FC < -fc_thres)&(deg1_filtered$pct.2>pct_thres))),]
deg2_filtered <- deg2_filtered[(((deg2_filtered$avg_log2FC > fc_thres)&(deg2_filtered$pct.1>pct_thres))|((deg2_filtered$avg_log2FC < -fc_thres)&(deg2_filtered$pct.2>pct_thres))),]
deg3_filtered <- deg3_filtered[(((deg3_filtered$avg_log2FC > fc_thres)&(deg3_filtered$pct.1>pct_thres))|((deg3_filtered$avg_log2FC < -fc_thres)&(deg3_filtered$pct.2>pct_thres))),]

```


```{r}
candidates_ec_deg <- intersect(intersect(row.names(deg1_filtered), row.names(deg2_filtered)), row.names(deg3_filtered))
candidates_ec_deg_passed_pos <- c()
candidates_ec_deg_passed_neg <- c()
candidates_ec_deg_passed_peak <- c()

diff <- 0
for (i in candidates_ec_deg){
  if (deg1_filtered[i, "avg_log2FC"] >0 & deg2_filtered[i, "avg_log2FC"] > 0 & deg3_filtered[i, "avg_log2FC"] > 0){
    if((deg1_filtered[i, "pct.1"] - deg1_filtered[i, "pct.2"] > diff) & (deg2_filtered[i, "pct.1"] - deg2_filtered[i, "pct.2"] > diff) & (deg3_filtered[i, "pct.1"] - deg3_filtered[i, "pct.2"] > diff)){
      candidates_ec_deg_passed_pos <- c(candidates_ec_deg_passed_pos, i)
    }
  }
  else if (deg1_filtered[i, "avg_log2FC"] <0 & deg2_filtered[i, "avg_log2FC"] < 0 & deg3_filtered[i, "avg_log2FC"] < 0){
    if((deg1_filtered[i, "pct.2"] - deg1_filtered[i, "pct.1"] > diff) & (deg2_filtered[i, "pct.2"] - deg2_filtered[i, "pct.1"] > diff) & (deg3_filtered[i, "pct.2"] - deg3_filtered[i, "pct.1"] > diff)){
      candidates_ec_deg_passed_neg <- c(candidates_ec_deg_passed_neg, i)
    }
  }
  else if (deg1_filtered[i, "avg_log2FC"] > 0 & deg2_filtered[i, "avg_log2FC"] > 0 & deg3_filtered[i, "avg_log2FC"] < 0){
    if((deg1_filtered[i, "pct.1"] - deg1_filtered[i, "pct.2"] > diff) & (deg2_filtered[i, "pct.1"] - deg2_filtered[i, "pct.2"] > diff) & (deg3_filtered[i, "pct.2"] - deg3_filtered[i, "pct.1"] > diff)){
      candidates_ec_deg_passed_peak <- c(candidates_ec_deg_passed_peak, i)
    }
  }
}
```

```{r}
print(candidates_ec_deg_passed_pos)
print(candidates_ec_deg_passed_neg)
```


```{r}
pdf(file= paste0(cwd, "/result/ec_up_genes.pdf"), onefile = TRUE)
for (i in seq(1, length(candidates_ec_deg_passed_pos), by = 4)){
  p <- VlnPlot(endo_cells, features = candidates_ec_deg_passed_pos[i:(i+3)], ncol = 2)
  print(p)
}
dev.off()


pdf(file= paste0(cwd, "/result/ec_down_genes.pdf"), onefile = TRUE)
for (i in seq(1, length(candidates_ec_deg_passed_neg), by = 4)){
  p <- VlnPlot(endo_cells, features = candidates_ec_deg_passed_neg[i:(i+3)], ncol = 2)
  print(p)
}
dev.off()


pdf(file= paste0(cwd, "/result/ec_peak_genes.pdf"), onefile = TRUE)
for (i in seq(1, length(candidates_ec_deg_passed_peak), by = 4)){
  p <- VlnPlot(endo_cells, features = candidates_ec_deg_passed_peak[i:(i+3)], ncol = 2)
  print(p)
}
dev.off()

```

```{r}
ec_markers <- ec_markers[ec_markers$pct.1 > 0.6 & ec_markers$pct.2 < 0.1 & ec_markers$p_val < 0.05, ]
write.csv(ec_markers, "result/deg_tab_ec_specific_46.csv")

# get 7xx degs -> 46 degs
```


```{r}
test <- subset(seu_data, idents = c("Endothelial"))
Idents(test) <- test@meta.data$characteristics..age

# RenameIdnets
test <- RenameIdents(test, 'E16.5' = "embryonic")
test <- RenameIdents(test, 'P0' = "postnatal", 'P5' = "postnatal")
test <- RenameIdents(test, 'P18' = "weaning",'P19' = "weaning",  'P23' = "weaning")
test <- RenameIdents(test, 'P120' = "adult", 'P132' = "adult")
Idents(test) <- factor(Idents(test), levels = c("embryonic", "postnatal", "weaning","adult"))
```


```{r}
DoHeatmap(object = test, features = c("Hipk3", "Bmpr2", "Mbnl1", "Pura"), group.by = "characteristics..age", label=FALSE) + theme(axis.text.x = element_text(size = 4), axis.text.y = element_text(size = 5), legend.text = element_text(size = 6)) + scale_fill_gradientn(colors = c("blue", "grey", "red"))


DoHeatmap(object = test, features = c("Ndufb5", "Cox7b", "Slc3a2", "Atp5b", "Car2" ,"Car4"), group.by = "characteristics..age", label=FALSE) + theme(axis.text.x = element_text(size = 4), axis.text.y = element_text(size = 5), legend.text = element_text(size = 6)) + scale_fill_gradientn(colors = c("blue", "grey", "red"))
DoHeatmap(object = test, features = candidates_ec_deg_passed_peak, group.by = "characteristics..age", label=FALSE) + theme(axis.text.x = element_text(size = 4), axis.text.y = element_text(size = 5), legend.text = element_text(size = 6)) + scale_fill_gradientn(colors = c("blue", "grey", "red"))


DoHeatmap(object = test, features = c("Slc3a2", "Car2", "Cox7b", "Atp5b", "Car4", "Dusp3", "Mbnl1", "Hipk3"), group.by = "characteristics..age", label=FALSE) + theme(axis.text.x = element_text(size = 4), axis.text.y = element_text(size = 5), legend.text = element_text(size = 6)) + scale_fill_gradientn(colors = c("yellow", "orange", "red"))
```




## ======================= biological sets =========================

```{r}

# Define the cell type vectors
brain_immunology <- c("PVM", "Endothelial", "VLMC", "Astro-adult", "OPC", "Cajal-Retzius", "Ependymal","MOL","NFOL", "Astro-adult", "Astro-juv", "Immature-Astro")
neurogenesis <- c("RGL", "RGL_young", "nIPC", "nIPC-perin", "Neuroblast", "Immature-GC", "Immature-Astro", "Astro-juv", "Immature-Pyr", "CA3-Pyr", "Ependymal",  "GC-juv","GC-adult", "GABA", "Immature-GABA", "Cajal-Retzius")
memory_circuit <- c("Cajal-Retzius", "CA3-Pyr", "Immature-Pyr", "GABA", "Astro-adult", "Immature-GABA", "Endothelial")
metabolic_system <- c("Astro-adult", "OPC", "MOL", "Endothelial", "Ependymal", "Immature-Astro", "VLMC", "Immature-GC")


# Subset the Seurat object based on each vector
brain_immunology_subset <- subset(seu_data, subset = characteristics..cell.cluster %in% brain_immunology)
neurogenesis_subset <- subset(seu_data, subset = characteristics..cell.cluster %in% neurogenesis)
memory_circuit_subset <- subset(seu_data, subset = characteristics..cell.cluster %in% memory_circuit)
metabolic_system_subset <- subset(seu_data, subset = characteristics..cell.cluster %in% metabolic_system)
```

```{r}
# Define a function to process each Seurat object and plot DimPlot
process_and_plot <- function(seurat_obj, title) {
  seurat_obj <- NormalizeData(seurat_obj)
  seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = 2000)
  seurat_obj <- ScaleData(seurat_obj)
  seurat_obj <- RunPCA(seurat_obj, verbose = FALSE)
  seurat_obj <- FindNeighbors(seurat_obj, dims = 1:30)
  seurat_obj <- FindClusters(seurat_obj, resolution = 1.0, verbose = FALSE)
  seurat_obj <- RunUMAP(seurat_obj, dims = 1:25)
  print(DimPlot(seurat_obj, reduction = "umap",  label = TRUE) + ggtitle(title))
  return(seurat_obj)
}

# 1. Brain Immunology Subset
brain_immunology_subset <- process_and_plot(brain_immunology_subset, "Brain Immunology Subset")
# 2. Neurogenesis Subset
neurogenesis_subset <- process_and_plot(neurogenesis_subset, "Neurogenesis Subset")
# 3. Memory Circuit Development and Pruning Subset
memory_circuit_subset <- process_and_plot(memory_circuit_subset, "Memory Circuit Development and Pruning Subset")
# 4. Metabolic System Subset
metabolic_system_subset <- process_and_plot(metabolic_system_subset, "Metabolic System Subset")

```

# plot dim plots for 4 sets - by clusters & cell types
```{r}
pdf("4sets_dim.pdf", width = 8, height = 7)
names <- c("brain immunology", "neurogenesis", "memory circuit", "metabolism")
name_idx <- 1
for (i in c(brain_immunology_subset, neurogenesis_subset, memory_circuit_subset, metabolic_system_subset))
{
  p <- DimPlot(i, reduction = "umap",  label = TRUE) + ggtitle(paste0(names[name_idx], " - clusters")) + theme(plot.title = element_text(hjust = 0.5))
  print(p)
  p <- DimPlot(i, group.by = "characteristics..cell.cluster", reduction = "umap",  label = TRUE) + ggtitle(paste0(names[name_idx], " - cell types")) + theme(plot.title = element_text(hjust = 0.5))
  print(p)
  
  name_idx <- name_idx+1
}
dev.off()
```


## ====> temporal-specific markers
```{r}
DefaultAssay(endo_cells)  <- "RNA"

deg1 <- FindMarkers(endo_cells, ident.1 = "postnatal", ident.2 = "embryonic", logfc.threshold = 0.2)
write.csv(deg1, "result/deg_tab_ec_postnatal.csv")
deg2 <- FindMarkers(endo_cells, ident.1 = "weaning", ident.2 = "postnatal", logfc.threshold = 0.2)
write.csv(deg2, "result/deg_tab_ec_weaning.csv")
deg3 <- FindMarkers(endo_cells, ident.1 = "adult", ident.2 = "weaning", logfc.threshold = 0.2)
write.csv(deg3, "result/deg_tab_ec_adult.csv")


names <- c("immunology", "neurogenesis", "memory", "metabolism")
name_idx <- 1
immunology_degs <- c()
neurogenesis_degs <- c()
memory_degs <- c()
metabolism_degs <- c()

temp_deg <- c()

for (i in c(brain_immunology_subset, neurogenesis_subset, memory_circuit_subset, metabolic_system_subset))
{
  DefaultAssay(i)  <- "RNA"
  i@meta.data$cluster <- i@active.ident
  Idents(i) <- i@meta.data$characteristics..age

  # RenameIdnets
  i <- RenameIdents(i, 'E16.5' = "embryonic")
  i <- RenameIdents(i, 'P0' = "postnatal", 'P5' = "postnatal")
  i <- RenameIdents(i, 'P18' = "weaning",'P19' = "weaning",  'P23' = "weaning")
  i <- RenameIdents(i, 'P120' = "adult", 'P132' = "adult")
  Idents(i) <- factor(Idents(i), levels = c("embryonic", "postnatal", "weaning","adult"))

  temp <- FindMarkers(i, ident.1 = "postnatal", ident.2 = "embryonic", logfc.threshold = 0.2)
  temp_deg <- c(temp_deg, temp)
  write.csv(temp, paste0("result/4sets/", names[name_idx], "postnatal_vs_embry.csv"))
  temp <- FindMarkers(i, ident.1 = "weaning", ident.2 = "postnatal", logfc.threshold = 0.2)
  temp_deg <- c(temp_deg, temp)
  write.csv(temp, paste0("result/4sets/", names[name_idx], "weaning_vs_postnatal.csv"))
  temp <- FindMarkers(i, ident.1 = "adult", ident.2 = "weaning", logfc.threshold = 0.2)
  temp_deg <- c(temp_deg, temp)
  write.csv(temp, paste0("result/4sets/", names[name_idx], "adult_vs_weaning.csv"))
  
  name_idx <- name_idx+1
}

```


## ======================= monocle3 =========================


```{r}
# Load Monocle3
library(monocle3)
library(SeuratWrappers)
```


```{r}
# Convert Seurat object to Monocle CDS (CellDataSet) for each subset
brain_immunology_cds <- as.cell_data_set(brain_immunology_subset)
neurogenesis_cds <- as.cell_data_set(neurogenesis_subset)
memory_circuit_cds <- as.cell_data_set(memory_circuit_subset)
metabolic_system_cds <- as.cell_data_set(metabolic_system_subset)
```


```{r}
# Preprocess and align cells along pseudotime for each subset
#brain_immunology_cds <- preprocess_cds(brain_immunology_cds, num_dim = 50)
#brain_immunology_cds <- reduce_dimension(brain_immunology_cds)
brain_immunology_cds <- cluster_cells(brain_immunology_cds, reduction_method = "UMAP")
brain_immunology_cds <- learn_graph(brain_immunology_cds)
brain_immunology_cds <- order_cells(brain_immunology_cds)

neurogenesis_cds <- cluster_cells(neurogenesis_cds, reduction_method = "UMAP")
neurogenesis_cds <- learn_graph(neurogenesis_cds)
neurogenesis_cds <- order_cells(neurogenesis_cds)


memory_circuit_cds <- cluster_cells(memory_circuit_cds, reduction_method = "UMAP")
memory_circuit_cds <- learn_graph(memory_circuit_cds)
memory_circuit_cds <- order_cells(memory_circuit_cds)

metabolic_system_cds <- cluster_cells(metabolic_system_cds, reduction_method = "UMAP")
metabolic_system_cds <- learn_graph(metabolic_system_cds)
metabolic_system_cds <- order_cells(metabolic_system_cds)
```


```{r}
# Plot pseudotime
pdf("traj_with_pseudotime.pdf", width = 8, height = 7)
print(plot_cells(brain_immunology_cds, color_cells_by = "pseudotime", show_trajectory_graph = TRUE) + ggtitle("immunology") + theme(plot.title = element_text(hjust = 0.5)))
print(plot_cells(neurogenesis_cds, color_cells_by = "pseudotime", show_trajectory_graph = TRUE) + ggtitle("neurogenesis") + theme(plot.title = element_text(hjust = 0.5)))
print(plot_cells(memory_circuit_cds, color_cells_by = "pseudotime", show_trajectory_graph = TRUE) + ggtitle("memory") + theme(plot.title = element_text(hjust = 0.5)))
print(plot_cells(metabolic_system_cds, color_cells_by = "pseudotime", show_trajectory_graph = TRUE) + ggtitle("metabolism") + theme(plot.title = element_text(hjust = 0.5)))
dev.off()



# Plot   others
plot_cells(brain_immunology_cds, show_trajectory_graph = TRUE)
plot_cells(brain_immunology_cds, color_cells_by = "partition", show_trajectory_graph = TRUE)
```

